---
title: "Calculating the final size of an epidemic"
output:
  bookdown::html_vignette2:
    fig_caption: yes
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Calculating the final size of an epidemic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

When considering responses to an epidemic, one of the first steps is to estimate how many infections there are likely to be over the course of the outbreak. This, the cumulative sum of all infections, is called the final size.
The final size of an epidemic can be calculated using methods detailed in @andreasen2011, @miller2012, @kucharski2014, and @bidari2016, and which are implemented in the _finalsize_ package.

::: {.alert .alert-primary}
## Use case {-}

An epidemic is underway. You want to know **how many individuals are likely to be infected in total**: this is the cumulative sum of all infections, or the _final size_ of the epidemic.
:::

::: {.alert .alert-secondary}
### What you have {-}

  1. An estimate of the infection's basic reproduction number $R_0$;
  2. An estimate of the distribution of the population in different demographic groups, often, age groups;
  3. An estimate of the contacts between individuals of different demographic groups; and
  4. An estimate of the susceptibility of each demographic group to the infection.

### What you assume {-}

  1. That the infection dynamics can be captured using a Susceptible-Infectious-Recovered (SIR) or Susceptible-Exposed-Infectious-Recovered (SEIR) model.
:::

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dev = "svglite"
)
```

## Getting _finalsize_

The _finalsize_ package is currently available on [Github](https://github.com/epiverse-trace/finalsize), and can be installed using the `remotes` or `devtools` packages.

```
# first install the remotes packages
install.packages("remotes")

# install finalsize using remotes
remotes::install_github("epiverse-trace/finalsize")
```

```{r}
library(finalsize)

# load useful packages
library(data.table)
library(ggplot2)
```

## Defining a value for $R_0$

A number of sophisticated statistical methods can be used to estimate the $R_0$ of an epidemic in its early stages. These are not covered here.

Instead, this example considers a infection with an $R_0$ of 2.0, similar to some strains of seasonal influenza.

```{r}
# define r0 as 2.0
r0 <- 2.0
```

## Getting and preparing contact and demography data

This example uses social contact data from the [POLYMOD](https://cordis.europa.eu/project/id/502084) project to estimate the final size of an epidemic in the U.K. These data are provided with the `socialmixr` package.

The contact data are divided into five age groups: 0 -- 5, 5 -- 18, 18 -- 40, 40 -- 65, and 65 and over, specified using the `age.limits` argument in `socialmixr::contact_matrix()`.
The `symmetric = TRUE` argument to `socialmixr::contact_matrix()` returns a symmetric contact matrix, so that the contacts reported by group $\{i\}$ of individuals from group $\{j\}$ are the same as those reported by group $\{j\}$ of group $\{i\}$.

The demographic data --- the number of individuals in each age group --- is also available through `socialmixr::contact_matrix()`.

```{r}
# get UK polymod data
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 5, 18, 40, 65),
  symmetric = TRUE
)

# get the contact matrix and demography data
contact_matrix <- t(contact_data$matrix)
demography_vector <- contact_data$demography$population
demography_data <- contact_data$demography

# scale the contact matrix so the largest eigenvalue is 1.0
contact_matrix <- contact_matrix / max(eigen(contact_matrix)$values)

# divide each row of the contact matrix by the corresponding demography
contact_matrix <- contact_matrix / demography_vector

n_demo_grps <- length(demography_vector)
```

::: {.alert .alert-secondary}

### Heterogeneous social mixing {-}

The social contact data used in this example illustrate an important way in which populations may be heterogeneous: the various age groups can have quite different patterns of social contacts.

The method implemented in _finalsize_ accounts for this heterogeneity. Other forms of population heterogeneity are covered in the ["Heterogeneous susceptibility"](varying_susceptibility.html) vignette.
:::

## Population susceptibility

In the early stages of an epidemic outbreak, a safe assumption might be that all age groups have a similar, high susceptibility to infection. This means that all individuals fall into a single category: highly or fully susceptible.

High uniform susceptibility can be modelled as a matrix with values of 1.0, with as many rows as there are demographic groups. The matrix has a single column, representing the single susceptibility group to which all individuals belong.

```{r}
# all individuals are equally and highly susceptible
n_susc_groups <- 1L
susc_guess <- 1.0
```

```{r}
susc_uniform <- matrix(
  data = susc_guess,
  nrow = n_demo_grps,
  ncol = n_susc_groups
)
```

Final size calculations also need to know the proportion of each demographic group $\{i\}$ that falls into the susceptibility group $\{j\}$. This distribution of age groups into susceptibility groups can be represented by the demography-susceptibility distribution matrix.

Since all individuals in each age group have the same susceptibility, there is no variation within age groups. Consequently, all individuals in each age group fall into the highly or fully susceptible group. This can be represented as a single-column matrix, with as many rows as age groups, and as many columns as susceptibility groups.

In this example, the matrix `p_susc_uniform` has `r n_demo_grps` rows, one for each age group, and only one column, for the single high susceptibility group that holds all individuals.

```{r}
p_susc_uniform <- matrix(
  data = 1.0,
  nrow = n_demo_grps,
  ncol = n_susc_groups
)
```

::: {.alert .alert-secondary}

#### Why susceptibility and p_susceptibility are matrices {-}

This example models susceptibility (`susc_uniform`) and the demography-in-susceptibility (`p_susc_uniform`) as matrices rather than vectors. This is because a single susceptibility group is a special case of the general final size equation.

_finalsize_ supports multiple susceptibility groups (this will be covered later), and these are more easily represented as a matrix, the _susceptibility matrix_.

Each element $\{i, j\}$ in this matrix represents the susceptibility of individuals in demographic group $\{i\}$, and susceptibility group $\{j\}$.

In this example, all individuals are equally susceptible to infection, and thus the susceptibility matrix (`susc_uniform`) has only a single column with identical values.

Consequently, the demography-susceptibility distribution matrix (`p_susc_uniform`) has the same dimensions, and all of its values are 1.0.

See the ["Heterogeneous susceptibility"](varying_susceptibility.html) example for more on cases where susceptibility varies within age groups.

:::

## Running `final_size`

The final size of the epidemic in the population can then be calculated using the only function in the package, `final_size`. This example allows the function to fall back on the default options for the arguments `solver` (`"iterative"`) and `control` (an empty list).

```{r}
# calculate final size
final_size_data <- final_size(
  r0 = r0,
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  susceptibility = susc_uniform,
  p_susceptibility = p_susc_uniform
)

# order demographic groups
final_size_data$demo_grp <- factor(
  final_size_data$demo_grp,
  levels = demography_data$age.group
)
```

### Visualise final sizes

```{r echo=FALSE, fig.cap ="Final size of an SIR epidemic in each age group. The final size is the cumulative number of infections in each age group over the course of the epidemic, expressed as a proportion of the respective age group.", fig.width=5, fig.height=4, class.source = 'fold-hide'}
# plot data
ggplot(final_size_data) +
  geom_col(
    aes(
      demo_grp, p_infected
    ),
    fill = "lightgrey",
    col = "black"
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.key.height = unit(2, "mm"),
    legend.title = ggtext::element_markdown(
      vjust = 1
    )
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  labs(
    x = "Age group",
    y = "% Infected"
  )
```

### Final size proportions to counts

`finalsize` returns the _proportion_ of each age (and risk) group infected in an epidemic outbreak. The final _counts_ of individuals infected can be visualised as well, by multiplying the final proportion of each age group infected with the total number of individuals in that group.

The example below show how this can be done.

```{r}
# prepare demography data
demography_data <- contact_data$demography

# merge final size counts with demography vector
final_size_data <- merge.data.table(
  final_size_data,
  demography_data,
  by.x = "demo_grp",
  by.y = "age.group"
)

# reset age group order
final_size_data$demo_grp <- factor(
  final_size_data$demo_grp,
  levels = contact_data$demography$age.group
)

# multiply counts with proportion infected
final_size_data$n_infected <- final_size_data$p_infected *
  final_size_data$population
```

```{r echo=FALSE, fig.cap="Final size of an epidemic outbreak in a population, for different values of infection $R_0$. Converting the final size proportions in each age group to counts shows that individuals aged 18 -- 65 make up the bulk of cases in most  outbreaks. This may be attributed to this being both the largest age range in the analysis (more years in this range than any other), and because more people fall into this wide range than others. Contrast this figure with the one above, in which similar _proportions_ of each age group are infected.", fig.width=5, fig.height=4, class.source = 'fold-hide'}
ggplot(final_size_data) +
  geom_col(
    aes(
      x = demo_grp, y = n_infected
    ),
    fill = "lightgrey",
    col = "black"
  ) +
  scale_y_continuous(
    labels = scales::comma_format(
      scale = 1e-6, suffix = "M"
    ),
    limits = c(0, 15e6)
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.key.height = unit(2, "mm"),
    legend.title = ggtext::element_markdown(
      vjust = 1
    )
  ) +
  coord_cartesian(
    expand = FALSE
  ) +
  labs(
    x = "Age group",
    y = "Number infected (millions)"
  )
```

## References
