---
title: "Examples of susceptiblity matrices"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Examples of susceptiblity matrices₀}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

_finalsize_ requires two matrices as arguments. The matrix format allows to easily represent __multiple__ demographic groups with __multiple__ susceptibility groups.

::: {.alert .alert-warning}
**New to _finalsize_?** It may help to read the ["Get started"](finalsize.html) vignette first!
:::

::: {.alert .alert-primary}
## Use case {-}

You need to create **two susceptibility matrices** to feed two arguments of the `final_size()` function. Here are five case examples with different scenarios for the content of each matrix.
:::

::: {.alert .alert-secondary}
### What we have {-}

  1. The susceptibility of a given number of demographic groups per a given number of susceptibility groups;
  2. The probability of the population in each demographic group and susceptibility group.

### What we assume {-}

  1. The complete partitioning of individuals into different demographic and infection risk groups (susceptibility groups).
:::


## _finalsize_ matrices

### Susceptibility matrix

Contains the _**susceptibility**_ of the population in each demographic group that belongs to each susceptibility group.

::: {.alert .alert-secondary}
$$
i \; demographic \; groups \left.\vphantom{\begin{matrix}
s_{1,1} & s_{1,2} & \cdots & s_{1,j} \\
s_{2,1} & s_{2,2} & \cdots & s_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
s_{i,1} & s_{i,2} & \cdots & s_{i,j} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} & s_{1,2} & \cdots & s_{1,j} \\
s_{2,1} & s_{2,2} & \cdots & s_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
s_{i,1} & s_{i,2} & \cdots & s_{i,j} 
\end{matrix}}^{\displaystyle j \; susceptibility \; groups}
\quad
s: susceptibility
$$

Each element $\{i, j\}$ in this matrix represents the susceptibility of individuals in demographic group $\{i\}$ (rows), and susceptibility group $\{j\}$ (columns).
:::

### Demography-susceptibility distribution matrix

Contains the _**probability**_ of the population in each demographic group that belongs to each susceptibility group.

::: {.alert .alert-secondary}
$$
i \; demographic \; groups \left.\vphantom{\begin{matrix}
p_{1,1} & p_{1,2} & \cdots & p_{1,j} \\
p_{2,1} & p_{2,2} & \cdots & p_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
p_{i,1} & p_{i,2} & \cdots & p_{i,j} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} & p_{1,2} & \cdots & p_{1,j} \\
p_{2,1} & p_{2,2} & \cdots & p_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
p_{i,1} & p_{i,2} & \cdots & p_{i,j} 
\end{matrix}}^{\displaystyle j \; susceptibility \; groups}
\quad
p: probability
$$

Each element $\{i, j\}$ in this matrix represents the proportion of each demographic group $\{i\}$ that falls into the susceptibility group $\{j\}$. Consequently, each of its **rows must always sum to 1.0**.
:::

Here we share five case examples with different scenarios for the content of each matrix.

## Homogenious susceptibility

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, message=FALSE, warning=FALSE, class.source = 'fold-hide'}
# load necessary packages
if (!require("dplyr")) install.packages("dplyr")
if (!require("tibble")) install.packages("tibble")

library(dplyr)
library(tibble)
```

### Case 1: Uniform susceptibility

In the population,
all individuals from all age groups
have 80% susceptibility to the infection

``` {r}
# susceptibility matrix
susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(0.8, 0.8, 0.8, 0.8, 0.8)) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

susceptibility
```

::: {.alert .alert-warning}
In this example, all individuals are equally susceptible to infection, and thus the **susceptibility matrix** (`susceptibility`) has only a single column with identical values.
:::

```{r}
# demography-susceptibility distribution matrix
p_susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0)) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
The __susceptibility matrix__ and the __demography-susceptibility distribution matrix__ has always the same dimensions.

$$
4 \; demographic \; groups \left.\vphantom{\begin{matrix}
s_{1,1} \\
s_{2,1} \\
s_{3,1} \\
s_{4,1} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} \\
s_{2,1} \\
s_{3,1} \\
s_{4,1} 
\end{matrix}}^{\displaystyle 1 \; susceptibility \; group}
\quad
4 \; demographic \; groups \left.\vphantom{\begin{matrix}
p_{1,1} \\
p_{2,1} \\
p_{3,1} \\
p_{4,1} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} \\
p_{2,1} \\
p_{3,1} \\
p_{4,1} 
\end{matrix}}^{\displaystyle 1 \; susceptibility \; group}
$$
:::

## Heterogeneous susceptibility

### Case 2: Between groups only

In the population,
there is different
susceptibility to the infection
between individuals of different age groups
from 20% (infants) to 100% (65+).

``` {r}
# susceptibility matrix
susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(0.2, 0.5, 0.6, 0.9, 1.0)) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0)) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

p_susceptibility
```

### Case 3: Within and between groups

In the population, there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%
to the 40% of each of the age groups.

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)) %>% 
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         immunised = c(0.4, 0.4, 0.4, 0.4, 0.4)) %>% 
  mutate(unimmunised = 1 - immunised) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
**Note** that because there are two susceptibility groups, the susceptibility matrix has two columns. The corresponding **demography-susceptibility distribution matrix** must also have two columns!

$$
4 \; demographic \; groups \left.\vphantom{\begin{matrix}
s_{1,1} & s_{1,2} \\
s_{2,1} & s_{2,2} \\
s_{3,1} & s_{3,2} \\
s_{4,1} & s_{4,2} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} & s_{1,2} \\
s_{2,1} & s_{2,2} \\
s_{3,1} & s_{3,2} \\
s_{4,1} & s_{4,2} 
\end{matrix}}^{\displaystyle 2 \; susceptibility \; groups}
\quad
4 \; demographic \; groups \left.\vphantom{\begin{matrix}
p_{1,1} & p_{1,2} \\
p_{2,1} & p_{2,2} \\
p_{3,1} & p_{3,2} \\
p_{4,1} & p_{4,2} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} & p_{1,2} \\
p_{2,1} & p_{2,2} \\
p_{3,1} & p_{3,2} \\
p_{4,1} & p_{4,2} 
\end{matrix}}^{\displaystyle 2 \; susceptibility \; groups}
$$
:::

### Case 4: Within groups in different proportion

In the population,
there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%.

Also the
immunization uptake rate
is in different proportions
for each of the age groups:
immunization increases with age
from 20% (infants) to 90% (65+)

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)) %>% 
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         immunised = c(0.2, 0.4, 0.6, 0.7, 0.9)) %>% 
  mutate(unimmunised = 1 - immunised) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

p_susceptibility
```

### Case 5: Within three groups

In the population,
there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%
to the 40% of each of the age groups.

Additionally,
10% of individuals
in each of the age groups
have
100% susceptibility,
due to no immunization
or not exposed
to similar pathogens previously.

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0),
         unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)) %>% 
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- 
  tibble(age_group = c("[0,5)","[5,18)","[18,40)","[40,65)","65+"),
         susceptible = c(0.1, 0.1, 0.1, 0.1, 0.1),
         immunised = c(0.4, 0.4, 0.4, 0.4, 0.4)) %>% 
  mutate(unimmunised = 1 - immunised - susceptible) %>% 
  column_to_rownames(var = "age_group") %>% 
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
**Recall** that each row of the demography-susceptibility distribution matrix must always sum to 1.0!
:::