---
title: "Guide to constructing susceptibility matrices"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
pkgdown:
  as_is: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Guide to constructing susceptibility matrices}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

Modelling the final size of an epidemic using _finalsize_ requires us to represent the susceptibility status of multiple demographic groups, as well as the proportions of each demographic group in potentially many susceptibility groups. This is done in final size calculations by representing the relevant system of differential equations [@miller2012] in matrix form. R's matrix class allows us to easily represent _multiple_ demographic groups with _multiple_ susceptibility groups.

This guide explains how to construct susceptibility (`susceptibility`) and demography-in-susceptibility (`p_susceptibility`) matrices to represent various scenarios that we might wish to model when estimating epidemic final sizes.

::: {.alert .alert-warning}
**New to _finalsize_?** It may help to read the ["Get started"](finalsize.html) vignette first!
:::

::: {.alert .alert-primary}
## Use case {-}

You need to create **two matrices**, `susceptibility` and `p_susceptibility`, to pass to the two arguments of the `final_size()` function. Here are five case examples with different scenarios for the content of each matrix.
:::

::: {.alert .alert-secondary}
### What we have {-}

  1. The susceptibility of a given number of demographic groups per a given number of susceptibility groups;
  2. The probability of the population in each demographic group and susceptibility group.

### What we assume {-}

  1. Complete partitioning of individuals into different demographic and susceptibility groups, i.e., that all individuals in the population fall into only one demographic group, and only one susceptibility group.
:::


## Primer on susceptibility matrices

### Susceptibility matrix

Represents the _**probability of infection**_ of each combination of demographic and infection-risk groups.

::: {.alert .alert-secondary}
$$
i \; demographic \; groups \left.\vphantom{\begin{matrix}
s_{1,1} & s_{1,2} & \cdots & s_{1,j} \\
s_{2,1} & s_{2,2} & \cdots & s_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
s_{i,1} & s_{i,2} & \cdots & s_{i,j} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} & s_{1,2} & \cdots & s_{1,j} \\
s_{2,1} & s_{2,2} & \cdots & s_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
s_{i,1} & s_{i,2} & \cdots & s_{i,j} 
\end{matrix}}^{\displaystyle j \; susceptibility \; groups}
\quad
s: susceptibility
$$

Each element $\{i, j\}$ in this matrix represents the susceptibility of individuals in demographic group $\{i\}$ (rows), and susceptibility group $\{j\}$ (columns).
:::

::: {.alert .alert-warning}
A common example of a demographic group in a population is an age group, and each population is typically made up of multiple age groups. 

Examples of infection-risk groups may include individuals of different immunization status, or with different prior exposure to pathogens. Final size calculations allow us to specify a different susceptibility to infection for each demography and risk group combination (e.g. fully vaccinated people over 65 years old).
:::

### Demography-in-susceptibility matrix

Represents the _**proportion (or probability)**_ that individuals in a demographic group are also in a specific susceptibility group.

::: {.alert .alert-secondary}
$$
i \; demographic \; groups \left.\vphantom{\begin{matrix}
p_{1,1} & p_{1,2} & \cdots & p_{1,j} \\
p_{2,1} & p_{2,2} & \cdots & p_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
p_{i,1} & p_{i,2} & \cdots & p_{i,j} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} & p_{1,2} & \cdots & p_{1,j} \\
p_{2,1} & p_{2,2} & \cdots & p_{2,j} \\
\vdots  & \vdots  & \ddots & \vdots  \\
p_{i,1} & p_{i,2} & \cdots & p_{i,j} 
\end{matrix}}^{\displaystyle j \; susceptibility \; groups}
\quad
p: probability
$$

Each element $\{i, j\}$ in this matrix represents the proportion of each demographic group $\{i\}$ that falls into the susceptibility group $\{j\}$. Consequently, each of its **rows must always sum to 1.0**.
:::

Here we share five case examples with different scenarios and how each influences the content and structure of the two matrices.

## Homogeneous susceptibility

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, message=FALSE, warning=FALSE, class.source = 'fold-hide'}
# load necessary packages
if (!require("dplyr")) install.packages("dplyr")
if (!require("tibble")) install.packages("tibble")

library(dplyr)
library(tibble)
```

### Case 1: Uniform susceptibility

This scenario assumes that all individuals from all age groups in the population of interest have 80% susceptibility to the infection.

``` {r}
# susceptibility matrix
susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(0.8, 0.8, 0.8, 0.8, 0.8)
  ) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility
```

::: {.alert .alert-warning}
In this example, all individuals and age groups are equally susceptible to infection; thus, the **susceptibility matrix** (`susceptibility`) has only a single column with identical values.
:::

```{r}
# demography-susceptibility distribution matrix
p_susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0)
  ) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
The __susceptibility matrix__ and the __demography-susceptibility distribution matrix__ must always have the same dimensions.

$$
4 \; demo. \; groups \left.\vphantom{\begin{matrix}
s_{1,1} \\
s_{2,1} \\
s_{3,1} \\
s_{4,1} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} \\
s_{2,1} \\
s_{3,1} \\
s_{4,1} 
\end{matrix}}^{\displaystyle 1 \; susceptibility \; group}
\quad
4 \; demo. \; groups \left.\vphantom{\begin{matrix}
p_{1,1} \\
p_{2,1} \\
p_{3,1} \\
p_{4,1} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} \\
p_{2,1} \\
p_{3,1} \\
p_{4,1} 
\end{matrix}}^{\displaystyle 1 \; susceptibility \; group}
$$
:::

## Heterogeneous susceptibility

### Case 2: Between groups only

In the population,
there is different
susceptibility to the infection
between individuals of different age groups
from 20% (infants) to 100% (65+).

``` {r}
# susceptibility matrix
susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(0.2, 0.5, 0.6, 0.9, 1.0)
  ) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0)
  ) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility
```

### Case 3: Within and between groups

In the population, there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%
to the 40% of each of the age groups.

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)
  ) %>%
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  immunised = c(0.4, 0.4, 0.4, 0.4, 0.4)
  ) %>%
  mutate(unimmunised = 1 - immunised) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
**Note** that because there are two susceptibility groups, the susceptibility matrix has two columns. The corresponding **demography-susceptibility distribution matrix** must also have two columns!

$$
4 \; demo. \; groups \left.\vphantom{\begin{matrix}
s_{1,1} & s_{1,2} \\
s_{2,1} & s_{2,2} \\
s_{3,1} & s_{3,2} \\
s_{4,1} & s_{4,2} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
s_{1,1} & s_{1,2} \\
s_{2,1} & s_{2,2} \\
s_{3,1} & s_{3,2} \\
s_{4,1} & s_{4,2} 
\end{matrix}}^{\displaystyle 2 \; susceptibility \; groups}
\quad
4 \; demo. \; groups \left.\vphantom{\begin{matrix}
p_{1,1} & p_{1,2} \\
p_{2,1} & p_{2,2} \\
p_{3,1} & p_{3,2} \\
p_{4,1} & p_{4,2} 
\end{matrix}}\right\{
\overbrace{
\begin{matrix}
p_{1,1} & p_{1,2} \\
p_{2,1} & p_{2,2} \\
p_{3,1} & p_{3,2} \\
p_{4,1} & p_{4,2} 
\end{matrix}}^{\displaystyle 2 \; susceptibility \; groups}
$$
:::

### Case 4: Within groups in different proportion

In the population,
there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%.

Also the
immunization uptake rate
is in different proportions
for each of the age groups:
immunization increases with age
from 20% (infants) to 90% (65+)

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)
  ) %>%
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  immunised = c(0.2, 0.4, 0.6, 0.7, 0.9)
  ) %>%
  mutate(unimmunised = 1 - immunised) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility
```

### Case 5: Within three groups

In the population,
there is different
susceptibility to the infection:

- between individuals of different age groups
from 20% (infants) to 100% (65+), and

- within individuals of the same age group
due the immunization effect of 25%
to the 40% of each of the age groups.

Additionally,
10% of individuals
in each of the age groups
have
100% susceptibility,
due to no immunization
or not exposed
to similar pathogens previously.

``` {r}
immunization_effect <- 0.25

# susceptibility matrix
susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(1.0, 1.0, 1.0, 1.0, 1.0),
  unimmunised = c(0.2, 0.5, 0.6, 0.9, 1.0)
  ) %>%
  mutate(immunised = unimmunised * (1 - immunization_effect)) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

susceptibility

# demography-susceptibility distribution matrix
p_susceptibility <- tibble(
  age_group = c("[0,5)", "[5,18)", "[18,40)", "[40,65)", "65+"),
  susceptible = c(0.1, 0.1, 0.1, 0.1, 0.1),
  immunised = c(0.4, 0.4, 0.4, 0.4, 0.4)
  ) %>%
  mutate(unimmunised = 1 - immunised - susceptible) %>%
  column_to_rownames(var = "age_group") %>%
  as.matrix()

p_susceptibility
```

::: {.alert .alert-warning}
**Recall** that each row of the demography-susceptibility distribution matrix must always sum to 1.0!
:::

## References
